Beyond Static Retrieval: Adaptive Method for Improving RAG Performance

### 2025-03-22 Paper/latex/paper_RAG.tex
update

\usepackage{graphicx}%
\usepackage{multirow}%
\usepackage{amsmath,amssymb,amsfonts}%
\usepackage{amsthm}%
\usepackage{mathrsfs}%
\usepackage[title]{appendix}%
\usepackage{xcolor}%
\usepackage{textcomp}%
\usepackage{manyfoot}%
\usepackage{booktabs}%
\usepackage{algorithm}%
\usepackage{algorithmicx}%
\usepackage{algpseudocode}%
\usepackage{listings}%

@@

\abstract{Retrieval-Augmented Generation (RAG) enhances language model responses by incorporating external knowledge. However, its effectiveness heavily depends on the quality of the retrieved documents. Using a fixed number of retrieved documents (k) may fail  to adapt to varying query complexity, often leading to suboptimal retrieval either including irrelevant documents or missing crucial ones. To address this issue, we propose a hybrid approach that dynamically adjusts K based on query characteristics while maintaining computational efficiency. Our method improves retrieval performance by maximizing relevant information and minimizing noise. Experimental results show that it outperforms fixed-K retrieval, offering a more effective solution for optimizing RAG systems. Future work will explore learning-based techniques to further refine dynamic K selection.}

@@

Large Language Models (LLMs)\cite{Radford2019} have emerged as state-of-the-art artificial intelligence systems that excel at understanding and generating human-like text,Built on deep learning architectures,particularly transformers \cite{vaswani2017attention},these models showcase impressive adaptability across diverse tasks, such as machine translation\cite{Gu2018SearchEngine}, text generation\cite{liang2024controllabletextgenerationlarge} and question answering\cite{Zhang2023}.Notable examples of LLMs include OpenAI GPT series\cite{brown2020language}, Google BERT\cite{devlin2018bert} and T5 \cite{raffel2020exploring}.Despite their impressive capabilities,LLMs suffer from hallucination\cite{ji2023survey}, often generating confident but incorrect information.This is a critical challenge for knowledge-intensive tasks requiring high accuracy\cite{kandpal2023large}.Additionally, their static nature being trained on fixed datasets\cite{bender2021dangers} limits their ability to update knowledge dynamically, making them prone to outdated or inconsistent responses, especially for time-sensitive information.\cite{Mousavi2025}.

Researchers have explored various approaches to address this limitation, including transfer learning\cite{zhuang2020comprehensive}a technique that improves pre-trained models performance in new tasks .Another technique is fine-tuning \cite{Howard2018} which adapt pre-trained models to specific tasks.Despite their strengths,they are computationally expensive and impractical for real-time knowledge updates. A promising alternative is Retrieval-Augmented Generation (RAG) \cite{lewis2020retrieval}.which integrates real-time retrieval from external knowledge sources with generative models enables LLMs to access up-to-date information without retraining,(RAG) faces several limitations. For instance,The performance of RAG heavily depends on the quality of the retriever \cite{karpukhin2020dpr}, as irrelevant or noisy documents can degrade the generator output. Additionally,A critical challenge is determining the optimal number of retrieved passages (K) to feed into the generator\cite{lewis2020retrieval}. often resulting in either insufficient or excessive information being retrieved, reduce coherence and increase computational overhead. This trade-off makes (k) a key factor in balancing performance and efficiency.

Current approaches, such as static K selection \cite{lewis2020retrieval}, dynamic K adjustment attempt to address this trade-off but often introduce complexity or latency,Hybrid approaches \cite{yuan2024hybrid},that integrates static retrieval with dynamic retrieval. However, these approaches face significant limitations, including , high computational overhead, and a dependency on high-quality metadata for optimal performance. These limitations highlight the need for a more robust approach to K selection, capable of balancing efficiency and accuracy across varying query types.

In this paper, we investigate the problem of selecting K in RAG and propose a novel approach to dynamically adjust K based on  thresholds and a Mixture of Logits (MoL)\cite{ding2024efficient} scoring mechanism, ensuring efficient and context-sensitive candidate selection while reducing computational costs. the use of  adaptive thresholds algorithm improves the relevance of retrieved documents, refining the set to include only the most contextually appropriate matches. Unlike traditional approaches, it adapts to the query and dataset characteristics, making it more flexible for diverse contexts. Furthermore, its design for handling large datasets ensures scalable and precise retrieval, which is crucial for RAG systems working with extensive corpora.

This paper contributes to the existing literature by reviewing the various approaches to Retrieval-Augmented Generation (RAG), with a specific focus on static, dynamic, and Hybrid selection strategies. Additionally, it provides a critical analysis of the strengths and limitations of these methods, as it identifies key gaps in their design and performance. On account of this, there arises a pressing need for an improved approach capable of addressing the challenges of both static and dynamic in retrieval-augmented systems. \\
The first section, Overview of Retrieval-Augmented Generation (RAG), aims to provide an a detailed explanation of the RAG framework, including its components and how it operates.\\
The following section,introduces the novel hybrid dynamic selection algorithm. Additionally, it provides detailed description of its methodology, explains its design, and discusses the manner in which it overcomes the limitations of existing  approaches. Particular emphasis is placed on the benefits of this approach, particularly in improving retrieval accuracy and computational efficiency.\\
The final section presents experimental results that demonstrate the effectiveness of the hybrid dynamic selection algorithm.highlight its superiority in terms of retrieval performance and adaptability to diverse query complexities.

\section{Related works}\label{sec2}
The selection of an optimal number of retrieved items (K) in Retrieval-Augmented Generation (RAG) systems plays a crucial role in ensuring accurate and efficient responses.Several methods, including static, dynamic, and hybrid approaches have been proposed to address this challenge While these methods show promise in improving performance,they often face trade-offs between adaptability, efficiency when applied to complex or diverse query patterns. In this section, we will explore these methods, highlight their respective strengths and weaknesses, and introduce our hybrid solution as a more robust and adaptable alternative.

Static Top-K selection in Retrieval :In traditional Retrieval-Augmented Generation (RAG) systems, static K-selection is a widely used method where a fixed number of top-k documents are retrieved based on their relevance scores. Sparse retrieval\cite{bai2020spartermlearningtermbasedsparse} methods where documents and queries are represented as high-dimensional, sparse vectors In these method , the "keys" represent documents to be retrieved, and the "values" are the importance or weight of those terms in the document such as query likelihood models\cite{Lafferty2001} ,TF-IDF\cite{Robertson1997} and BM25\cite{Zaragoza2009} often rely on this static approach to identify and rank documents. For instance, if k=5, the system retrieves the top 5 documents deemed most relevant to a query, regardless of the query's complexity or the variability in relevance distribution among documents.
A major drawback of this approach is its rigidity. Since the number of retrieved documents (k) is fixed, it fails to adapt to the varying complexity of queries.This one-size-fits-all strategy can lead to suboptimal performance across diverse datasets or tasks..

Dynamic top-k Selection in Retrieval :addresses the limitations of static k by varying the number of retrieved documents based on the specific query or context. These approaches typically involve heuristics, machine learning models, or adaptive algorithms to determine the optimal K for each query such as Dynamic Query Cutoff (DQC)\cite{Culpepper2016} This approach uses machine learning models trained on query-document features to predict the optimal (k) for retrieval, A reinforcement learning-based approach\cite{Zhou2021} where the agent learns to adjust (k) by maximizing downstream task performance (e.g accurate response generation in RAG) Another example is the Dynamic Selection of k Nearest Neighbors \cite{DynamicKNN2012} where the selection process often depends on the distribution of neighbors, their distances to the query point, or their relevance scores.
However, these methods come with limitations.They often introduce computational overhead due to the need for dynamic evaluation, which can increase latency in large-scale systems. Additionally, their performance heavily depends on the quality of the underlying models or distance metrics, making them sensitive to noisy data. As a result, achieving generalization across diverse datasets or domains remains a significant challenge.

Hybrid Approaches For Top-K selection in Retrieval: combine elements of static and dynamic K-selection to achieve a balance between simplicity and adaptability.These systems might use a static k as a baseline but dynamically adjust it such as ,the "Blended RAG"\cite{sawarkar2024blendedragimprovingrag} approach enhances retrieval accuracy by combining semantic search techniques with hybrid query strategies,Furthermore STAYKATE (Static-Dynamic Hybrid Selection)\cite{Zhu2024} enhances LLM performance in scientific information extraction by integrating representativeness sampling from active learning with a retrieval-based strategy. Another example is DR-RAG (Dynamic Relevance for Retrieval-Augmented Generation)\cite{hei2024drragapplyingdynamicdocument} is a two-stage retrieval framework.first, it retrieves an initial set of (k1) documents followed by a classifier that evaluates the dynamic relevance between documents to enhance recall and retrieve additional relevant 
(k2) documents, however hybrid methods may face challenges in determining when to apply static or dynamic k-selection effectively, leading to potential trade-offs between retrieval accuracy and system efficiency. These limitations highlight the ongoing need for robust mechanisms that dynamically and efficiently adapt k to the unique characteristics of each query and dataset.

@@

\begin{figure}[h]
	\centering
	\includegraphics[width=0.9\textwidth]{rag.JPG}
	\caption{RAG system component\cite{sharma2024rag}. }\label{rag.PNG}
\end{figure}
As illustrated in figure \ref{rag.PNG}, the RAG system is built around two main components:the retriever and the generator. The retriever's role is to find relevant information from a pre-built data store, while the generator utilizes this retrieved data to produce coherent and contextually appropriate content. The overall RAG process operates as follows :
\subsection{Retrieval in RAG Systems}\label{subsec2}

@@

\subsection{Generator  in RAG Systems}\label{subsec2}
In Retrieval-Augmented Generation (RAG) systems, the generator\cite{huang2024surveyretrievalaugmentedtextgeneration} is a key component responsible for crafting the final output. It works by combining the retrieved information with the users input query to produce a well-structured and contextually relevant response .

@@

In various generative tasks, different models are selected based on the specific requirements: Models like BART\cite{lewis2019bart}, GPT\cite{brown2020language}and and T5 \cite{raffel2020exploring} are commonly used for tasks such as text summarization and translation ,Vision-Language Models (VLMs) \cite{radford2021learningtransferablevisualmodels} are employed to generate textual descriptions from images , for text-to-Code Generation: Models like OpenAIs Codex \cite{chen2021evaluatinglargelanguagemodels} are designed to translate natural language prompts into executable code.

@@

\noindent
2. \textbf{Dynamic Threshold Adjustment}:  
To improve retrieval quality, \textbf{Mixture of Logits (MoL)} scores are computed for each candidate \(x \in G\). The adaptive gating weights \(\pi_p(q, x)\) allow the algorithm to dynamically adjust the retrieval threshold \(T_{\text{adaptive}}\) based on the MoL scores. The scoring function is defined as:
\begin{equation}
	\phi(q, x) = \sum_{p=1}^{P} \pi_p(q, x) \cdot \langle f_p(q), g_p(x) \rangle.
\end{equation}
The adaptive threshold \(T_{\text{adaptive}}\) is set as the minimum score among the candidates:
\begin{equation}
	T_{\text{adaptive}} = \min\{s \mid s \in G\}.
\end{equation}

\noindent
3. \textbf{Refinement and Top-K Selection}:  
Using the adaptive threshold \(T_{\text{adaptive}}\), additional relevant candidates are retrieved, expanding the candidate set \(G'\). This is achieved by including candidates whose scores exceed the threshold:
\begin{equation}
	G' \leftarrow G \cup \{x \mid s_p \geq T_{\text{adaptive}}\}.
\end{equation}
The algorithm then sorts \(G'\) based on MoL scores to select the most relevant top-k candidates.

\noindent
4. \textbf{Exact Top-K Selection}:  
Finally, the candidates in \(G'\) are sorted by their MoL scores, and the exact top-k items are extracted:
\begin{equation}
	G_{\text{final}} = \text{Top-k}(G', \phi(q, x)).
\end{equation}
As shown in Algorithm\ref{ALgo1}, the retrieval process dynamically adjusts the threshold based on MoL scores.
\begin{algorithm}
\caption{Hybrid Exact Top-k with Threshold-Based k Selection}
\label{ALgo1}
	\begin{algorithmic}[1]
     \State \textbf{Input:} Query $q$, Set of items $X$, Component-level embeddings: $f_p(q)$, $g_p(x)$ for $p \in P$, $x \in X$, Initial threshold $T_{\text{init}}$ 
    \State \textbf{Output:} Exact top $k$ items based on dynamic threshold   selection, $G_{\text{final}}$

@@

	    \textbf{1. Initial Candidate Retrieval:}
		\For{each component $p \in P$}
		\State Compute dot product scores: 
		\[
		S_p \gets  \{ \langle f_p(q), g_p(x) \rangle : x \in X_p \}
		\]
		\State Retrieve items with scores $S_p \geq T_{\text{init}}$
		\State Add these items to $G$
		\EndFor

		 \textbf{2. Adjust k Dynamically:}
		\For{each $x \in G$}
		\State Compute MoL scores $s \gets \phi(q, x)$ where :
		\[
		\phi(q, x) \gets \sum_{p=1}^{P} \pi_p(q, x) \cdot \langle f_p(q), g_p(x) \rangle
		\]
		\State Set $T_{\text{adaptive}} = \min \{s : s \in G\}$
		\EndFor

@@

		 \textbf{4. Select Exact Top-k Items:}
		 \For{each component $p \in P$}
		\State Compute MoL scores for all items in $G'$
		\State Sort $G'$ by MoL scores in descending order
		\State Select the top $k$ items from $G'$ where $k$ is the number of items in $G'$ exceeding $T_{\text{adaptive}}$
			\EndFor
		\State\textbf{ Return:} $G_{\text{final}}$ \Comment{Retrieve Top $k$ items from $G'$}
	\end{algorithmic}
\end{algorithm}
\newpage
\section{Experimental Results}\label{sec3}
We evaluate the performance of the retrieval algorithm and the Mixture-of-Logits (MoL) approach on the next-item prediction task in recommendation systems\cite{Zhu_2018}\cite{kang2018selfattentivesequentialrecommendation}. We conducted experiments using two MovieLens datasets: ML-100K and ML-1M \cite{Harper2015}  which are widely used benchmarks for evaluating sequential recommendation models.\\
\textbf{MovieLens-100K} \\
100,000 ratings (1-5 scale) from 943 users on 1,682 movies\\
Each user has rated at least 20 movies \\
Includes user demographic information (age, gender, occupation, zip code)\\
\textbf{MovieLens-1M} \\
1,000,209 ratings from 6,040 users on approximately 3,900 movies \\
Data collected from users who joined MovieLens in 2000 \\
Represents a larger-scale recommendation scenario \\
\subsection{ Setup }
For both datasets, We utilize the SASRec architecture as our sequential user encoder, a model renowned for achieving state-of-the-art performance in next-item prediction tasks. This architecture processes the user's historical interaction sequence, generating embeddings that encapsulate the user's preferences at each time step. These embeddings serve as the foundation for predicting the next item in the sequence.

The query \( q \) represents the user's state at a specific time step, derived from their interaction history. In the MoL (Mixture of Logits) framework,\( q \)is transformed into \( Pq \)
embeddings through a multi-layer perceptron (MLP).
\subsection{Hyperparameter settings.}
For fair comparison, we maintained consistent architectural choices and training conditions across all experiments , We conducted an extensive hyperparameter analysis comparing both approaches (Hybrid+SAS and MoL+SAS) across different architectural configurations.All experiments were Implemented in TensorFlow and trained on a Google Colab environment with a T4 GPU. We use the Adam optimizer with a learning rate of 0.001. For the hybrid algorithm, we initialize the threshold (Tinit) to 0.3 and adaptively adjust it during training.We discuss detailed hyperparameter settings in table \ref{tab:100k_movies} ,table \ref{tab:1m_movies}

\begin{table}[ht]
	\centering
	\large
	\renewcommand{\arraystretch}{1.5} % Increase row spacing
	\resizebox{\textwidth}{!}{  % Scale table to fit width
		\begin{tabular}{|l|c|c|c|c|c|c|c|c|}
			\hline
			\textbf{Model} & \textbf{Max Sequence Length} & \textbf{Embedding Dimension} & \textbf{Number of Heads} & \textbf{Feedforward Dimension} & \textbf{Batch Size} & \textbf{Epochs} & \textbf{Val Loss} & \textbf{Val Accuracy} \\ 
			\hline
			Hybrid+SAS & 50  & 128  & 2 & 128  & 128 & 12 & 5.3036 & 0.1584  \\ \hline
			MoL+SAS    & 50  & 128  & 2 & 128  & 128 & 12 & 5.3152 & 0.1582  \\ \hline
			Hybrid+SAS & 128 & 256  & 4 & 256  & 128 & 10 & 3.6364 & 0.4301  \\ \hline
			MoL+SAS    & 128 & 256  & 4 & 256  & 128 & 10 & 3.6374 & 0.4299 \\\hline
			Hybrid+SAS & 512 & 512  & 4 & 512  & 128 & 10 & 1.2808 & 0.8120 \\ \hline
			MoL+SAS    & 512 & 512  & 4 & 512  & 128 & 10 & 1.3050 & 0.8016 \\\hline
   \end{tabular}
}
\caption{Results on 100kMovies dataset}
\label{tab:100k_movies}
\end{table}

\begin{table}[ht]
	\centering
	\large
	\renewcommand{\arraystretch}{1.5} % Increase row spacing
	\resizebox{\textwidth}{!}{  % Scale table to fit width
		\begin{tabular}{|l|c|c|c|c|c|c|c|c|}
			\hline
		\textbf{Model} & \textbf{Max Sequence Length} & \textbf{Embedding Dimension} & \textbf{Number of Heads} & \textbf{Feedforward Dimension} & \textbf{Batch Size} & \textbf{Epochs} & \textbf{Val Loss} & \textbf{Val Accuracy} \\ 
		\hline
			Hybrid+SAS & 50  & 128  & 2 & 128  & 128 & 10 & 4.5721 & 0.1530  \\ \hline
			MoL+SAS    & 50  & 128  & 2 & 128  & 128 & 10 & 4.5733 & 0.1526  \\ \hline
			Hybrid+SAS & 128 & 256  & 4 & 256  & 128 & 10 & 3.4152 & 0.3680  \\ \hline
			MoL+SAS    & 128 & 256  & 4 & 256  & 128 & 10 & 3.4671 & 0.3627  \\ \hline
			Hybrid+SAS & 512 & 512  & 4 & 512  & 128 & 10 & 1.0275 & 0.8350  \\ \hline
			MoL+SAS    & 512 & 512  & 4 & 512  & 128 & 10 & 1.0350 & 0.8276  \\ \hline

		\end{tabular}
	}
	\caption{Results on 1M Movies dataset}
	\label{tab:1m_movies}
\end{table}
\newpage
Both approaches demonstrate significant performance improvements as model capacity increases, with larger configurations consistently delivering better results. For instance, when increasing the model's capacity such as expanding the (Max Sequence Length: 512, Embedding Dimension: 512, Number of Heads: 4, Feed-Forward Dimension: 512) the hybrid algorithm combined with SASRec (Hybrid+SAS) achieves notable gains, particularly in the most resource-intensive setup. On the ML-100K dataset, Hybrid+SAS reaches a score of\textbf{ 0.8120} compared to the baseline's \textbf{0.8016}, while on ML-1M, it achieves \textbf{0.8350} versus \textbf{0.8276.} The ML-1M dataset generally benefits more from increased capacity, with the performance gap between datasets narrowing as the model scales. While smaller configurations provide a balance of efficiency and performance, the largest configuration, despite its higher computational demands, yields the best results, making it suitable for scenarios where resources are not a constraint. Both approaches show similar benefits from scaling, but Hybrid+SAS maintains a consistent edge in performance.
\subsection{Evaluation metrics}

We employ standard ranking metrics widely used in sequential recommendation:

>>>>

\usepackage{graphicx}%
\usepackage{multirow}%
\usepackage{amsmath,amssymb,amsfonts}%
\usepackage{amsthm}%
\usepackage{mathrsfs}%
\usepackage[title]{appendix}%
\usepackage{xcolor}%
\usepackage{textcomp}%
\usepackage{manyfoot}%
\usepackage{booktabs}%
\usepackage{algorithm}%
\usepackage{algorithmicx}%
\usepackage{algpseudocode}%
\usepackage{listings}%
\usepackage{subcaption}

@@

\abstract{Retrieval-Augmented Generation (RAG) enhances language model responses by incorporating external knowledge. However, its effectiveness heavily depends on the quality of the retrieved documents. Using a fixed number of retrieved documents K may fail  to adapt to varying query complexity, often leading to suboptimal retrieval either including irrelevant documents or missing crucial ones. To address this issue, we propose a hybrid method that dynamically adjusts K based on query characteristics while maintaining computational efficiency. Our method improves retrieval performance by maximizing relevant information and minimizing noise. Experimental results show that it outperforms fixed-K retrieval, offering a more effective solution for optimizing RAG systems. Compared to traditional methods, it dynamically adjusts based on the query and dataset characteristics, offering greater adaptability across various contexts.}

@@

Large Language Models (LLMs) have emerged as state-of-the-art artificial intelligence systems that excel at understanding and generating human-like text\cite{Radford2019}, Built on deep learning architectures, particularly transformers \cite{vaswani2017attention},these models showcase impressive adaptability across diverse tasks, such as machine translation\cite{Gu2018SearchEngine}, text generation\cite{liang2024controllabletextgenerationlarge} and question answering\cite{Zhang2023}. Notable examples of LLMs include OpenAI GPT series\cite{brown2020language}, Google BERT\cite{devlin2018bert} and T5 \cite{raffel2020exploring}. Despite their impressive capabilities, LLMs suffer from hallucination\cite{ji2023survey}, often generating confident but incorrect information. This is a critical challenge for knowledge-intensive tasks requiring high accuracy\cite{kandpal2023large}. Additionally, their static nature being trained on fixed datasets\cite{bender2021dangers} limits their ability to update knowledge dynamically, making them prone to outdated or inconsistent responses, especially for time-sensitive information.\cite{Mousavi2025}.

Researchers have explored various approaches to address this limitation, including transfer learning\cite{zhuang2020comprehensive}, fine-tuning \cite{Howard2018} which adapt pre-trained models to specific tasks. Despite their strengths, they are computationally expensive and impractical for real-time knowledge updates. A promising alternative is Retrieval-Augmented Generation (RAG) \cite{lewis2020retrieval}, which integrates real-time retrieval from external knowledge sources with generative models, enables LLMs to access up-to-date information without retraining,(RAG) faces several limitations. For instance, The performance of RAG heavily depends on the quality of the retriever \cite{karpukhin2020dpr}, as irrelevant or noisy documents can degrade the generator output. Additionally, a critical challenge is determining the optimal number of retrieved passages K to feed into the generator\cite{lewis2020retrieval}. Often resulting in either insufficient or excessive information being retrieved, reduce coherence and increase computational overhead. This trade-off makes K a key factor in balancing performance and efficiency.

Current approaches, such as static K selection \cite{lewis2020retrieval}, dynamic K adjustment attempt to address this trade-off but often introduce complexity or latency, Hybrid approaches \cite{yuan2024hybrid},integrates static retrieval with dynamic retrieval. However, these approaches face significant limitations, including, high computational overhead, and a dependency on high-quality metadata for optimal performance. These limitations highlight the need for a more K selection robust approach, capable of balancing efficiency and accuracy across varying query types.

In this paper, we investigate the problem of selecting K in RAG and propose a novel approach to dynamically adjust K based on  thresholds and a Mixture of Logits (MoL)\cite{ding2024efficient} scoring mechanism, ensuring efficient and context-sensitive candidate selection while reducing computational costs. The use of  adaptive thresholds algorithm improves the relevance of retrieved documents, refining the set to include only the most contextually appropriate matches. Unlike traditional approaches, it adapts to the query and dataset characteristics, making it more flexible for diverse contexts. Furthermore, its design for handling large datasets ensures scalable and precise retrieval, which is crucial for RAG systems working with extensive corpora.

The remainder of this paper is as follows :We start by presenting an overview of the research problem,  We summarize existing solutions, and highlight the gaps that our work addresses through a more dynamic approach to k selection.\\
Section 2: summarizes existing solutions, additionally, it provides a critical analysis of the strengths and limitations of these methods, as it identifies key gaps in their design and performance. As a result, there is a critical need for a more effective approach.\\
Section 3: aims to provide a detailed explanation of the RAG framework, including its components and how it operates.\\
Section 4: introduces the novel hybrid dynamic selection algorithm. Additionally, it provides detailed description of its methodology, explains its design, and discusses the manner in which it overcomes the limitations of existing  approaches. Particular emphasis is placed on the benefits of this approach, particularly in improving retrieval accuracy and computational efficiency.\\
Section 5: presents experimental results that demonstrate the effectiveness of the hybrid dynamic selection algorithm. Highlight its superiority in terms of retrieval performance and adaptability to diverse query complexities.

\section{Related works}\label{sec2}
The selection of an optimal number of retrieved items K in Retrieval-Augmented Generation (RAG) systems plays a crucial role in ensuring accurate and efficient responses. Several methods, including static, dynamic, and hybrid approaches have been proposed to address this challenge While these methods show promise in improving performance, they often face trade-offs between adaptability, efficiency when applied to complex or diverse query patterns. In this section, we will explore these methods, highlight their respective strengths and weaknesses, and introduce our hybrid solution as a more robust and adaptable alternative.

Static Top-K selection in Retrieval :In traditional Retrieval-Augmented Generation (RAG) systems, static K-selection is a widely used method where a fixed number of top-k documents are retrieved based on their relevance scores. Sparse retrieval\cite{bai2020spartermlearningtermbasedsparse} methods where documents and queries are represented as high-dimensional, sparse vectors In these method, the "keys" represent documents to be retrieved, and the "values" are the importance or weight of those terms in the document such as query likelihood models\cite{Lafferty2001}, TF-IDF\cite{Robertson1997} and BM25\cite{Zaragoza2009} often rely on this static approach to identify and rank documents. For instance, if k=5, the system retrieves the top 5 documents deemed most relevant to a query, regardless of the query's complexity or the variability in relevance distribution among documents.
A major drawback of this approach is its rigidity. Since the number of retrieved documents K is fixed, it fails to adapt to the varying complexity of queries. This one-size-fits-all strategy can lead to suboptimal performance across diverse datasets or tasks.

Dynamic top-K Selection in Retrieval :addresses the limitations of static K by varying the number of retrieved documents based on the specific query or context. These approaches typically involve heuristics, machine learning models, or adaptive algorithms to determine the optimal K for each query such as Dynamic Query Cutoff (DQC)\cite{Culpepper2016} This approach uses machine learning models trained on query-document features to predict the optimal K for retrieval, a reinforcement learning-based approach\cite{Zhou2021} where the agent learns to adjust K by maximizing downstream task performance (e.g accurate response generation in RAG) Another example is the Dynamic Selection of K Nearest Neighbors \cite{DynamicKNN2012} where the selection process often depends on the distribution of neighbors, their distances to the query point, or their relevance scores.
However, these methods come with limitations. They often introduce computational overhead due to the need for dynamic evaluation, which can increase latency in large-scale systems. Additionally, their performance heavily depends on the quality of the underlying models or distance metrics, making them sensitive to noisy data. As a result, achieving generalization across diverse datasets or domains remains a significant challenge.

Hybrid Approaches For Top-K selection in Retrieval: combine elements of static and dynamic K-selection to achieve a balance between simplicity and adaptability. These systems might use a static K as a baseline but dynamically adjust it such as, the "Blended RAG"\cite{sawarkar2024blendedragimprovingrag} approach enhances retrieval accuracy by combining semantic search techniques with hybrid query strategies, Furthermore STAYKATE (Static-Dynamic Hybrid Selection)\cite{Zhu2024} enhances LLM performance in scientific information extraction by integrating representativeness sampling from active learning with a retrieval-based strategy. Another example is DR-RAG (Dynamic Relevance for Retrieval-Augmented Generation)\cite{hei2024drragapplyingdynamicdocument} First, it retrieves an initial set of $k_1$ documents. Then, a classifier dynamically evaluates their relevance and retrieves additional relevant $K_2$ documents to enhance recall. However hybrid methods may face challenges in determining when to apply static or dynamic K-selection effectively, leading to potential trade-offs between retrieval accuracy and system efficiency. These limitations highlight the ongoing need for robust mechanisms that dynamically and efficiently adapt K to the unique characteristics of each query and dataset.

@@

\begin{figure}[h]
	\centering
	\includegraphics[width=0.9\textwidth]{rag.JPG}
	\caption{RAG system component\cite{sharma2024rag}. }\label{rag.PNG}
\end{figure}
As illustrated in figure \ref{rag.PNG}, the RAG system is built around two main components: the retriever and the generator. The retriever's role is to find relevant information from a pre-built data store, while the generator utilizes this retrieved data to produce coherent and contextually appropriate content. The overall RAG process operates as follows :
\subsection{Retrieval in RAG Systems}\label{subsec2}

@@

\subsection{Generator  in RAG Systems}\label{subsec2}
In Retrieval-Augmented Generation (RAG) systems, the generator\cite{huang2024surveyretrievalaugmentedtextgeneration} is a key component responsible for crafting the final output. It works by combining the retrieved information with the users input query to produce a well-structured and contextually relevant response.

@@

In various generative tasks, different models are selected based on the specific requirements: Models like BART\cite{lewis2019bart}, GPT\cite{brown2020language}and and T5 \cite{raffel2020exploring} are commonly used for tasks such as text summarization and translation, Vision-Language Models (VLMs) \cite{radford2021learningtransferablevisualmodels} are employed to generate textual descriptions from images, for text-to-Code Generation: Models like OpenAIs Codex \cite{chen2021evaluatinglargelanguagemodels} are designed to translate natural language prompts into executable code.

@@

\noindent
2.\textbf{ Initial Candidate Retrieval:} This step involves computing similarity scores between a query representation and item representations for each feature component $p \in P$.

\begin{equation}
	S_p \gets  \{ \langle f_p(q), g_p(x) \rangle : x \in X_p \}
	\label{eq:initial_retrieval}
\end{equation}
Here, $f_p(q)$ and $g_p(x)$ represent the feature embeddings of the query $q$ and item $x$ for the $p$-th component, respectively. The dot product $\langle f_p(q), g_p(x) \rangle$ measures the relevance of each item $x \in X_p$ with respect to $q$, producing a set of scores $S_p$ .

\noindent
3. \textbf{Dynamic Threshold Adjustment}:  
To improve retrieval quality, \textbf{Mixture of Logits (MoL)} scores are computed for each candidate \(x \in G\). The adaptive gating weights \(\pi_p(q, x)\) allow the algorithm to dynamically adjust the retrieval threshold \(T_{\text{adaptive}}\) based on the MoL scores. The scoring function is defined as:
\begin{equation}
	\phi(q, x) = \sum_{p=1}^{P} \pi_p(q, x) \cdot \langle f_p(q), g_p(x) \rangle.
	\label{similarity_function}
\end{equation}
The adaptive threshold \(T_{\text{adaptive}}\) is set as the minimum score among the candidates:
\begin{equation}
	T_{\text{adaptive}} = \min\{s \mid s \in G\}.
\end{equation}

\noindent
4. \textbf{Refinement and Top-K Selection}:  
Using the adaptive threshold \(T_{\text{adaptive}}\), additional relevant candidates are retrieved, expanding the candidate set \(G'\). This is achieved by including candidates whose scores exceed the threshold:
\begin{equation}
	G' \leftarrow G \cup \{x \mid s_p \geq T_{\text{adaptive}}\}.
\end{equation}
The algorithm then sorts \(G'\) based on MoL scores to select the most relevant top-k candidates.

\noindent
4. \textbf{Exact Top-K Selection}:  
Finally, the candidates in \(G'\) are sorted by their MoL scores, and the exact top-k items are extracted:
\begin{equation}
	G_{\text{final}} = \text{Top-k}(G', \phi(q, x)).
\end{equation}
As shown in Algorithm\ref{ALgo1}, the retrieval process dynamically adjusts the threshold based on MoL scores.
\begin{algorithm}
\caption{Hybrid Exact Top-k with Threshold-Based k Selection}
\label{ALgo1}
 \textbf{Input:} Query $q$, Set of items $X$, Component-level embeddings: $f_p(q)$, $g_p(x)$ for $p \in P$, $x \in X$, Initial threshold $T_{\text{init}}$ \\
 \textbf{Output:} Exact top $k$ items based on dynamic threshold   selection, $G_{\text{final}}$

	\begin{algorithmic}[1]

@@

	    \textbf{1. Initial Candidate Retrieval:}
		\For{each component $p \in P$}
		\State Compute dot product scores:  with (eq\eqref{eq:initial_retrieval})

		\State Retrieve items with scores $S_p \geq T_{\text{init}}$
		\State Add these items to $G$
		\EndFor

		 \textbf{2. Adjust k Dynamically:}
		\For{each $x \in G$}
		\State Compute MoL scores $s \gets \phi(q, x)$ using : eq\eqref{similarity_function}
		\State Set $T_{\text{adaptive}} = \min \{s : s \in G\}$
		\EndFor

@@

		 \textbf{4. Select Exact Top-k Items:}
		 \For{each component $p \in P$}
		\State Compute MoL scores for all items in $G'$
		\State Sort $G'$ by MoL scores in descending order
		\State Select the top $k$ items from $G'$ where $k$ is the number of items in $G'$ exceeding $T_{\text{adaptive}}$
			\EndFor
		\State\textbf{ Return:} $G_{\text{final}}$ \Comment{Retrieve Top $k$ items from $G'$}
	\end{algorithmic}
\end{algorithm}
\newpage
\subsection{Evaluation metrics in recommendation system}

The retrieval algorithm is integrated into a recommendation system, and its effectiveness is assessed using standard ranking metrics. These metrics are widely used in sequential recommendation tasks to evaluate the model's ability to rank relevant items higher in a user's preference list:
